---
title: "Build R packages using GitHub Actions"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Build R packages using GitHub Actions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Using the set of GitHub Actions provided by [r-wasm/actions](https://github.com/r-wasm/actions), it is possible to build a list of Wasm R packages and automatically deploy the resulting package repository or library image to a GitHub Pages URL. This workflow simplifies the process of deploying a set of R packages for use with webR, and enables continuous integration.

## Setting up the GitHub repository

First, create a new GitHub repository [following GitHub's instructions](https://docs.github.com/en/get-started/quickstart/create-a-repo#) to initialise a new empty git repo.

Create a file named `packages`, containing a list of [R package references](https://r-lib.github.io/pkgdepends/reference/pkg_refs.html). Add one R package per line, and custom R packages hosted on GitHub may also be included here. For example:

```
cli
dplyr
tidyverse/ggplot2@v3.4.4
```

Next, create a new GitHub Actions workflow file at `.github/workflows/deploy.yml`, by running

```r
usethis::use_github_action(
  url = "https://raw.githubusercontent.com/r-wasm/actions/v1/examples/deploy-cran-repo.yml",
  save_as = "deploy.yml"
)
```

The workflow contents will have two workflow jobs. The first job builds the list of R packages into a package repository and uploads it as an artifact file. The second job downloads and deploys the package repository to GitHub Pages.

The workflow file should look like this:

```yaml
# Workflow derived from https://github.com/r-wasm/actions/tree/v1/examples
# Need help debugging build failures? Start at https://github.com/r-lib/actions#where-to-find-help
on:
  push:
    # Only build on main or master branch
    branches: [main, master]
  # Or when triggered manually
  workflow_dispatch:

name: Build and deploy wasm R package repository

jobs:
  # Reads `./packages` for package references to put
  # into a CRAN-like repository hosted on GitHub pages
  deploy-cran-repo:
    uses: r-wasm/actions/.github/workflows/deploy-cran-repo.yml@v1
    permissions:
      # To download GitHub Packages within action
      repository-projects: read
      # For publishing to pages environment
      pages: write
      id-token: write
```

Commit the new GitHub Actions file changes, then push the commit to GitHub.


## The GitHub Actions build process

The GitHub Actions workflow should automatically start to run and build your list of packages. You should be able to see progress of the website build step in the **Actions** section of your GitHub project.

After a little while, your GitHub Pages website will be ready and webR should be able to install your package from a GitHub Pages URL:

```{r eval=FALSE}
webr::install("cli", repos = "http://username.github.io/my-wasm-repo/")
#> Downloading webR package: cli
```

Further usage details can be found in the [r-wasm/actions GitHub documentation](https://github.com/r-wasm/actions/tree/v1/.github/workflows/README.md).

## Using an R package library image

An Emscripten filesystem image containing an R package library may also be built and attached to a GitHub package release. If you'd prefer to mount an R package library, rather than install packages from a CRAN-like repo, use the `release-file-system-image.yml` workflow.

```r
usethis::use_github_action(
  url = "https://raw.githubusercontent.com/r-wasm/actions/v1/examples/release-file-system-image.yml"
)
```

Commit the new GitHub Actions file, then make a release through the GitHub web interface. GitHub Actions will build a Wasm filesystem image for your package and its dependencies and upload it as asset files for that specific package release.

Once the Github Action has finished, the `library.data` and `library.js.metadata` assets files can be downloaded from the GitHub releases page. The filesystem image files should then be made available through static hosting in some way, for example though further GitHub Actions steps to upload the filesystem image files as part of a custom GitHub Pages website.^[In principle, it is possible to directly mount the filesystem image from the release asset URL using `webr::mount()`. However, in web browsers this will likely be blocked due to the [CORS](https://developer.mozilla.org/en-US/docs/Web/HTTP/CORS) mechanism.]

Finally, in webR, mount the statically hosted filesystem image and set the `.libPaths()` to load a package from the package library:

```{r eval=FALSE}
webr::mount("/my-library", "https://org.github.io/repo/download/library.data")
.libPaths(c(.libPaths(), "/my-library"))
library(dplyr)
#> Attaching package: ‘dplyr’
#>
#> The following objects are masked from ‘package:stats’:
#>
#>     filter, lag
#>
#> The following objects are masked from ‘package:base’:
#>
#>     intersect, setdiff, setequal, union
```

Further information about Emscripten filesystem images can be found in the `vignette("mount-fs-image.Rmd")` article.
