---
title: "Build R packages using GitHub Actions"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Build R packages using GitHub Actions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Using the set of GitHub Actions provided by [r-wasm/actions](https://github.com/r-wasm/actions), it is possible to build a list of Wasm R packages and automatically deploy the resulting package repository or library image to a GitHub Pages URL. This workflow simplifies the process of deploying a set of R packages for use with webR, and enables continuous integration.

## Setting up the GitHub repository

First, create a new GitHub repository [following GitHub's instructions](https://docs.github.com/en/get-started/quickstart/create-a-repo#) to intialise a new empty git repo.

Create a file named `packages`, containing a list of [R package references](https://r-lib.github.io/pkgdepends/reference/pkg_refs.html). Add one R package per line, and custom R packages hosted on GitHub may also be included here. For example:

```
cli
dplyr
tidyverse/ggplot2@v3.4.4
```

Next, create a new GitHub Actions workflow file at `.github/workflows/deploy.yml`, by running

```r
usethis::use_github_action(
  url = "https://raw.githubusercontent.com/r-wasm/actions/v1/examples/deploy-wasm-repo.yml",
  save_as = "deploy.yml"
)
```

The workflow contents will have two workflow jobs. The first job builds the list of R packages into a package repository and uploads it as an artifact file. The second job downloads and deploys the package repository to GitHub Pages.

The workflow file should look like this:

```yaml
on:
  push:
    branches: [main, master]
  pull_request:
    branches: [main, master]
  workflow_dispatch:

name: Build wasm R package repository

jobs:
  build:
    uses: r-wasm/actions/.github/workflows/build.yml@v1
  deploy:
    name: Deploy to GitHub pages
    needs: build
    runs-on: ubuntu-latest
    permissions:
      pages: write
      id-token: write
    environment:
      name: github-pages
      url: ${{ steps.deployment.outputs.page_url }}
    steps:
      - uses: actions/checkout@v3
      - name: Download wasm artifacts
        uses: r-wasm/actions/download-wasm-artifacts@v1
        with:
          repo-path: _site
      - name: Upload Pages artifact
        uses: actions/upload-pages-artifact@v2
      - name: Deploy to GitHub Pages
        id: deployment
        uses: actions/deploy-pages@v2
```

Commit the new GitHub Actions file changes, then push the commit to GitHub.


## The GitHub Actions build process

The GitHub Actions workflow should automatically start to run and build your list of packages. You should be able to see progress of the website build step in the **Actions** section of your GitHub project.

After a little while, your GitHub Pages website will be ready and webR should be able to install your package from a GitHub Pages URL:

```{r eval=FALSE}
webr::install("cli", repos = "http://username.github.io/my-wasm-repo/")
#> Downloading webR package: cli
```

Further usage details for `r-wasm/actions/build-wasm-packages` can be found in the [GitHub documentation](https://github.com/r-wasm/actions/tree/v1/build-wasm-packages).

## Using an R package library image

An Emscripten filesystem image containing an R package library may also be built and uploaded to GitHub Pages. If you'd prefer to mount an R package library, rather than install packages from a CRAN-like repo, use the `image-path` parameter to prepare the Emscripten filesystem image for upload to your GitHub Pages site:

```yaml
- name: Download wasm artifacts
  uses: r-wasm/actions/download-wasm-artifacts@v1
  with:
    image-path: _site
```

Then, in webR, mount the filesystem image and set the `.libPaths()` to load a package from the package library:

```{r eval=FALSE}
webr::mount("/my-library", "http://username.github.io/my-wasm-repo/library.data")
.libPaths(c(.libPaths(), "/my-library"))
library(dplyr)
#> Attaching package: ‘dplyr’
#>
#> The following objects are masked from ‘package:stats’:
#>
#>     filter, lag
#>
#> The following objects are masked from ‘package:base’:
#>
#>     intersect, setdiff, setequal, union
```

Further information about Emscripten filesystem images can be found in the `vignette("mount-fs-image.Rmd")` article.
