---
title: "Build R packages using GitHub Actions"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Build R packages using GitHub Actions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Using the set of GitHub Actions provided by [r-wasm/actions](https://github.com/r-wasm/actions), it is possible to build a list of Wasm R packages and automatically deploy the resulting package repository or library image to a GitHub Pages URL. This workflow simplifies the process of deploying a set of R packages for use with webR, and enables continuous integration.

## Setting up the GitHub repository

First, create a new GitHub repository [following GitHub's instructions](https://docs.github.com/en/get-started/quickstart/create-a-repo#) to intialise a new empty git repo.

Create a file named `packages`, containing a list of [R package references](https://r-lib.github.io/pkgdepends/reference/pkg_refs.html). Add one R package per line, and custom R packages hosted on GitHub may also be included here. For example:

```
cli
dplyr
tidyverse/ggplot2@v3.4.4
```

Next, create a new GitHub Actions workflow file at `.github/workflows/build.yml`, with contents:

```yaml
on:
  push:
    branches: [main, master]
  pull_request:
    branches: [main, master]
  workflow_dispatch:

name: Build wasm R package repository

jobs:
  build:
    runs-on: ubuntu-latest
    container: ghcr.io/r-wasm/webr:main
    steps:
    - uses: actions/checkout@v3
    - name: Build wasm packages
      uses: r-wasm/actions/build-wasm-packages@v1
  deploy:
    name: Deploy to GitHub pages
    needs: build
    runs-on: ubuntu-latest
    permissions:
      pages: write
      id-token: write
    environment:
      name: github-pages
      url: ${{ steps.deployment.outputs.page_url }}
    steps:
      - uses: actions/checkout@v3
      - name: Download wasm R package repository artifact
        uses: actions/download-artifact@v3
        with:
          name: wasm-repo
          path: _site
      - name: Upload Pages artifact
        uses: actions/upload-pages-artifact@v2
      - name: Deploy to GitHub Pages
        uses: actions/deploy-pages@v2
```

Commit the changes above, then push your repository to GitHub.


## The GitHub Actions build process

The GitHub Actions workflow should automatically start to run and build your list of packages. You should be able to see progress of the website build step in the **Actions** section of your GitHub project.

After a little while, your GitHub Pages website will be ready and webR should be able to install your package from a GitHub Pages URL:

```{r eval=FALSE}
webr::install("cli", repos = "http://username.github.io/my-wasm-repo/")
#> Downloading webR package: cli
```

Further usage details for `r-wasm/actions/build-wasm-packages` can be found in the [GitHub documentation](https://github.com/r-wasm/actions/tree/v1/build-wasm-packages).

## Using an R package library image

An Emscripten filesystem image containing an R package library may also be built and uploaded to GitHub pages. If you'd prefer to mount a package library image, rather than install R packages from a CRAN-like repo, upload the `wasm-image` artifact to your GitHub Pages site:

```yaml
- name: Download wasm R package repository artifact
  uses: actions/download-artifact@v3
  with:
    name: wasm-image
    path: _site
```

Then, in webR, mount the filesystem image and set the `.libPaths()` to load a package from the package library:

```{r eval=FALSE}
webr::mount("/my-library", "http://username.github.io/my-wasm-repo/library.data")
.libPaths(c(.libPaths(), "/my-library"))
library(dplyr)
#> Attaching package: ‘dplyr’
#> 
#> The following objects are masked from ‘package:stats’:
#> 
#>     filter, lag
#> 
#> The following objects are masked from ‘package:base’:
#> 
#>     intersect, setdiff, setequal, union
```

Further information about Emscripten filesystem images can be found in the `vignette("mount-fs-image.Rmd")` article.
