---
title: "Build R packages using GitHub Actions"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Build R packages using GitHub Actions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

Using the set of GitHub Actions provided by [r-wasm/actions](https://github.com/r-wasm/actions), it is possible to automatically build and deploy WebAssembly binary versions of R packages. This workflow simplifies the process of deploying R packages for use with webR, and facilitates continuous integration.

## Deploying an R package on release

It is possible to build an R package for WebAssembly, optionally including other R package dependencies in the output, using the `release-file-system-image.yml` GitHub Actions workflow.

First, add the GitHub action to your R package repository:

```r
usethis::use_github_action(
  url = "https://raw.githubusercontent.com/r-wasm/actions/v1/examples/release-file-system-image.yml"
)
```

Commit the new GitHub Actions file, then make a package release through the GitHub web interface. GitHub Actions will build a Wasm filesystem image for your package and its dependencies and upload it as asset files for that specific package release.

Once the Github Action has finished running, the asset files `library.data` and `library.js.metadata` can be downloaded from the GitHub releases page.

If you are using your Wasm R package in a Shiny app with `shinylive::export()`, it will now be downloaded automatically as part of the app export. Otherwise, read on for details on how to make the resulting R package binaries available to a webR application.

### Hosting the resulting package

In principle, it is possible to directly mount the filesystem image from the release asset URL using `webr::mount()`. However, in web browsers this will likely be blocked due to the [CORS](https://developer.mozilla.org/en-US/docs/Web/HTTP/CORS) mechanism. Instead, the files should be uploaded to static hosting in some way, for example by additional GitHub Actions steps that uploads the release assets to GitHub Pages.

An example GitHub action that uploads release assets to GitHub pages, in addition to pkgdown documentation, can be found [here](https://github.com/r-wasm/actions/blob/main/examples/rwasm-binary-and-pkgdown-site.yml).

Once the files have been hosted on GitHub Pages, or otherwise, mount the filesystem image in webR and set the `.libPaths()` to load R packages from the package library:

```{r eval=FALSE}
webr::mount("/my-library", "https://org.github.io/repo/download/library.data")
.libPaths(c(.libPaths(), "/my-library"))
library(dplyr)
#> Attaching package: ‘dplyr’
#>
#> The following objects are masked from ‘package:stats’:
#>
#>     filter, lag
#>
#> The following objects are masked from ‘package:base’:
#>
#>     intersect, setdiff, setequal, union
```

Further information about Emscripten filesystem images can be found in the `vignette("mount-fs-image.Rmd")` article.

## Creating a WebAssembly CRAN-like repository

Alternatively, if you have multiple R packages you may want to build a custom WebAssembly CRAN-like repository on GitHub Pages. This workflow is useful if you want to manage several packages and do not mind creating a new repository, separate from your R package source, to do so.

First, create a new GitHub repository [following GitHub's instructions](https://docs.github.com/en/get-started/quickstart/create-a-repo#) to initialise a new empty git repo.

Create a file named `packages`, containing a list of [R package references](https://r-lib.github.io/pkgdepends/reference/pkg_refs.html). Add one R package per line, and custom R packages hosted on GitHub may also be included here. For example:

```
cli
dplyr
tidyverse/ggplot2@v3.4.4
```

Next, create a new GitHub Actions workflow file at `.github/workflows/deploy.yml`, by running

```r
usethis::use_github_action(
  url = "https://raw.githubusercontent.com/r-wasm/actions/v1/examples/deploy-cran-repo.yml",
  save_as = "deploy.yml"
)
```

The workflow contents will have two workflow jobs. The first job builds the list of R packages into a package repository and uploads it as an artifact file. The second job downloads and deploys the package repository to GitHub Pages.

The workflow file should look like this:

```yaml
# Workflow derived from https://github.com/r-wasm/actions/tree/v1/examples
# Need help debugging build failures? Start at https://github.com/r-lib/actions#where-to-find-help
on:
  push:
    # Only build on main or master branch
    branches: [main, master]
  # Or when triggered manually
  workflow_dispatch:

name: Build and deploy wasm R package repository

jobs:
  # Reads `./packages` for package references to put
  # into a CRAN-like repository hosted on GitHub pages
  deploy-cran-repo:
    uses: r-wasm/actions/.github/workflows/deploy-cran-repo.yml@v1
    permissions:
      # To download GitHub Packages within action
      repository-projects: read
      # For publishing to pages environment
      pages: write
      id-token: write
```

Commit the new GitHub Actions file changes, then push the commit to GitHub.

### The GitHub Actions build process

The GitHub Actions workflow should automatically start to run and build your list of packages. You should be able to see progress of the website build step in the **Actions** section of your GitHub project.

After a little while, your GitHub Pages website will be ready and webR should be able to install your package from a GitHub Pages URL:

```{r eval=FALSE}
webr::install("cli", repos = "http://username.github.io/my-wasm-repo/")
#> Downloading webR package: cli
```

Further usage details can be found in the [r-wasm/actions GitHub documentation](https://github.com/r-wasm/actions/blob/v1/.github/workflows/README.md).
