---
title: "Mounting filesystem images"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Mounting filesystem images}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The Emscripten WebAssembly (Wasm) environment provides a virtual filesystem (VFS) which supports the concept of *mounting*. With this, an entire file and directory structure can be packaged into a filesystem image, efficiently making individual files or entire R package libraries available for use in webR.

## Create filesystem images

### Emscripten's `file_packager` tool

The [`file_packager`](https://emscripten.org/docs/porting/files/packaging_files.html#packaging-using-the-file-packager-tool) tool, provided by Emscripten, takes in a directory structure as input and produces a webR compatible filesystem image as output. The [`file_packager`](https://emscripten.org/docs/porting/files/packaging_files.html#packaging-using-the-file-packager-tool) tool may be invoked from the [rwasm](https://r-wasm.github.io/rwasm/) package:

```{r eval=FALSE}
rwasm::file_packager("./input", out_dir = ".", out_name = "output")
```

It can also be invoked directly using its CLI^[See the [`file_packager`](https://emscripten.org/docs/porting/files/packaging_files.html#packaging-using-the-file-packager-tool) Emscripten documentation for details.], if you prefer:

```bash
$ file_packager output.data --preload ./input@/ \
    --separate-metadata --js-output=output.js
```

In the above examples, the files in the directory `./input` are packaged and an output filesystem image is created^[When using the `file_packager` CLI, a third file named `output.js` will also be created. If you only plan to mount the image using webR, this file may be discarded.] consisting of a data file, `output.data`, and a metadata file, `output.js.metadata`.

To prepare for mounting the filesystem image with webR, ensure that both files have the same basename (in this example, `output`). The resulting URLs or relative paths for the two files should differ only by the file extension.

#### Compression

Filesystem image `.data` files may optionally be `gzip` compressed prior to deployment. The file extension for compressed filesystem images should be `.data.gz`, and compression should be indicated by setting the property `gzip: true` on the metadata JSON stored in the `.js.metadata` file.

**NOTE**: Loading compressed VFS images requires at least version 0.4.1 of webR.

### Mount `.tar` archives as a filesystem image

Archives in `.tar` format, optionally gzip compressed as `.tar.gz` or `.tgz` files, can also be used as filesystem images by pre-processing the `.tar` archive using the `rwasm::add_tar_index()` function. The function reads archive contents and appends the required filesystem metadata to the end of the `.tar` archive data in a way that is understood by webR. For further information about the format see the [Technical details for .tar archive metadata](tar-metadata.html) article.

```{r eval=FALSE}
rwasm::add_tar_index("./path/to/archive.tar.gz")
# Appending virtual filesystem metadata for: ./path/to/archive.tar.gz
```

Once processed by `rwasm::add_tar_index()`, the `.tar` archive can be deployed and used directly as a filesystem image.

## Mounting filesystem images

When running in a web browser, the [`webr::mount()`](https://docs.r-wasm.org/webr/latest/api/r.html#mount) function downloads and mounts a filesystem image from a URL source, using the `WORKERFS` filesystem type.

```{r eval=FALSE}
webr::mount(
  mountpoint = "/data",
  source = "https://example.com/output.data"
)
```

Filesystem images should be deployed to static file hosting^[e.g. GitHub Pages, Netlify, AWS S3, etc.] and the resulting URL provided as the source argument. The image will be mounted in the virtual filesystem under the path given by the `mountpoint` argument. If the `mountpoint` directory does not exist, it will be created prior to mounting.

When running under Node.js, the source may also be provided as a relative path to a filesystem image on disk.

To test filesystem images before deployment, serve them using a local static webserver. See the Local Testing section below for an example using `httpuv::runStaticServer()` in R.

## Building an R package library image

A collection of R packages can be collected and bundled into a single filesystem image for mounting. 

To build an R package library image we must first build one or more Wasm R packages using `add_pkg()`. As an example, let's build a package with a few hard dependencies. Ensure that you are running R in an environment with access to Wasm development tools^[See the "Setting up the WebAssembly toolchain" section in `vignette("rwasm")` for further details.], then run:

```{r eval=FALSE}
rwasm::add_pkg("dplyr")
```

After the build process has completed, the new `repo` directory contains a CRAN-like package repository with R packages build for Wasm.

Next, run the following to build an Emscripten VFS image:

```{r eval=FALSE}
rwasm::make_vfs_library()
```

By default, this function will create a new directory named `vfs` if it does not exist. The files `vfs/library.data` and `vfs/library.js.metadata` together form an Emscripten filesystem image containing an R package library consisting of all the packages previously added to the CRAN-like repository in `repo` using `add_pkg()`.

### Local testing

The following R command starts a local web server to serve your filesystem image for testing^[Ensure that the latest version of the `httpuv` package is installed so that the `?httpuv::runStaticServer` function is available.]. When serving your files locally, be sure to include the `Access-Control-Allow-Origin: *` HTTP header, required for downloading files from a cross-origin server by the [CORS](https://developer.mozilla.org/en-US/docs/Web/HTTP/CORS) mechanism.

```r
httpuv::runStaticServer(
  dir = ".",
  port = 9090,
  browse = FALSE,
  headers = list("Access-Control-Allow-Origin" =  "*")
)
```

Once the web server is running start a webR session in your browser, such as the console at <https://webr.r-wasm.org/latest/>. Use `webr::mount()` to make the R library image available somewhere on the VFS^[You might need to adjust the path portion of the URL, depending on your set-up. If it does not work, you could also try `"http://127.0.0.1:9090"` or `"http://127.0.0.1:9090/output/vfs"`]:

```{r eval=FALSE}
webr::mount("/my-library", "http://127.0.0.1:9090/vfs/library.data")
```

Once mounted, the contents of the filesystem image are available at `/my-library` in the virtual filesystem.

```{r eval=FALSE}
list.files("/my-library")
#>  [1] "R6"         "cli"        "dplyr"     "fansi"      "generics"   "glue"
#>  [7] "lifecycle"  "magrittr"   "pillar"    "pkgconfig"  "rlang"      "tibble"
#> [13] "tidyselect" "utf8"       "vctrs"     "withr"
```

This new directory should be added to R's `.libPaths()`, so that R packages may be loaded from the new library.

```{r eval=FALSE}
.libPaths(c(.libPaths(), "/my-library"))
library(dplyr)
#> Attaching package: ‘dplyr’
#>
#> The following objects are masked from ‘package:stats’:
#>
#>     filter, lag
#>
#> The following objects are masked from ‘package:base’:
#>
#>     intersect, setdiff, setequal, union
```
